Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2025-06-25 paper.qmd
Add and updated some words with Paul's comment

This is a benchmark single-cell data set collected on Human Peripheral Blood Mononuclear Cells (PBMC3k) as used in @pbmc2019. Single-cell data measures the gene expression of individual cells in a sample of tissue (see  for example, @haque2017). This type of data is used to obtain an understanding of cellular level behavior and heterogeneity in their activity. Clustering of single-cell data is used to identify groups of cells with similar expression profiles. NLDR is often used to summarize the cluster structure. Usually, NLDR does not use the cluster labels to compute the layout, but uses color to represent the cluster labels when it is plotted.

In this data there are $2622$ single cells and $1000$ gene expressions (variables). Following their pre-processing, different NLDR techniques were performed on the first nine principal components. @fig-NLDR-variety shows this data using a variety of methods, and different hyper-parameters. You can see that the result is wildly different depending on the choices. Layout a is a reproduction of the layout published in @chen2024. This layout suggests that the data has three very well separated clusters, each with an odd shape. The question is whether this accurately represents the cluster structure in the data, or whether they should have chosen b or c or d or e or f or g. This is what our new method can help with -- to decide which is the more accurate \gD{} representation of the cluster structure in the \pD{} data.

>>>>

This is a benchmark single-cell RNA-Seq data set collected on Human Peripheral Blood Mononuclear Cells (PBMC3k) as used in @pbmc2019. Single-cell data measures the gene expression of individual cells in a sample of tissue (see  for example, @haque2017). This type of data is used to obtain an understanding of cellular level behavior and heterogeneity in their activity. Clustering of single-cell data is used to identify groups of cells with similar expression profiles. NLDR is often used to summarize the cluster structure. Usually, NLDR does not use the cluster labels to compute the layout, but uses color to represent the cluster labels when it is plotted.

In this data there are $2622$ single cells and $1000$ gene expressions (variables). Following their pre-processing, different NLDR techniques were performed on the first nine principal components. @fig-NLDR-variety shows this data using a variety of methods, and different hyper-parameters. You can see that the result is wildly different depending on the choices. Layout a is a reproduction of the layout published in @chen2024. This layout suggests that the data has three very well separated clusters, each with an odd shape. The question is whether this accurately represents the cluster structure in the data, or whether they should have chosen b or c or d or e or f or g or h. This is what our new method can help with -- to decide which is the more accurate \gD{} representation of the cluster structure in the \pD{} data.
